---
title: "Data Manipulation with tidyverse"
subtitle: "Data Manipulation  — 2025"
author: "Christophe Bontemps (SIAP)"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: kate
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    df_print: paged
    code_folding: show
---

```{css, echo=FALSE}
/* ── General ── */
body { font-family: 'DM Sans', 'Segoe UI', sans-serif; font-size: 15px; color: #1a1916; }
h1, h2, h3 { font-weight: 600; }
h2 { border-bottom: 2px solid #e2dfd8; padding-bottom: 6px; margin-top: 2rem; }
h3 { color: #2a6496; margin-top: 1.6rem; }

/* ── Exercise boxes ── */
.exercise {
  background: #f7f6f2;
  border-left: 4px solid #2a6496;
  border-radius: 0 8px 8px 0;
  padding: 16px 20px;
  margin: 1.2rem 0;
}
.exercise .ex-label {
  font-size: 11px;
  font-weight: 600;
  text-transform: uppercase;
  letter-spacing: .12em;
  color: #2a6496;
  margin-bottom: 6px;
}
.exercise p { margin: 0; color: #3a3a38; line-height: 1.7; }
.exercise strong { color: #8a5c0a; }
.exercise code { background: #e8e6e0; padding: 1px 5px; border-radius: 3px; font-size: 13px; }

/* ── Hint boxes ── */
.hint {
  background: #eaf1f8;
  border: 1px solid #b5d4f4;
  border-radius: 8px;
  padding: 12px 16px;
  margin: .8rem 0;
  font-size: 13.5px;
  color: #185fa5;
  line-height: 1.6;
}
.hint::before { content: "💡 Hint: "; font-weight: 600; }

/* ── Solution boxes ── */
.solution {
  background: #f0ecfb;
  border: 1px solid #c4b5f8;
  border-radius: 8px;
  padding: 12px 16px;
  margin: .8rem 0;
  font-size: 13.5px;
  color: #5b3fa6;
}
.solution::before { content: "✅ Solution — expand code block above"; font-weight: 600; display: block; margin-bottom: 4px; }

/* ── Schema cards ── */
.schema-wrap { display: flex; gap: 14px; flex-wrap: wrap; margin: 1rem 0 1.4rem; }
.schema-card {
  background: #fff;
  border: 1px solid #e2dfd8;
  border-radius: 10px;
  padding: 12px 16px;
  min-width: 200px;
  font-size: 13px;
}
.schema-card .sc-title {
  font-family: monospace;
  font-size: 12px;
  color: #1a7070;
  font-weight: 600;
  margin-bottom: 8px;
  display: flex; align-items: center; gap: 6px;
}
.schema-card .sc-title::before { content: ''; display: inline-block; width: 7px; height: 7px; border-radius: 50%; background: #1a7070; }
.schema-card table { width: 100%; border-collapse: collapse; }
.schema-card td { padding: 2px 6px 2px 0; font-family: monospace; font-size: 12px; }
.schema-card td:first-child { color: #1a1916; font-weight: 500; }
.schema-card td:last-child { color: #9b9790; }
.key-col td:first-child { color: #8a5c0a !important; }

/* ── SQL comparison block ── */
.sql-compare {
  display: flex; gap: 12px; flex-wrap: wrap; margin: 1rem 0;
}
.sql-box, .r-box {
  flex: 1; min-width: 220px;
  border-radius: 8px;
  padding: 12px 14px;
  font-family: monospace;
  font-size: 12.5px;
  line-height: 1.8;
}
.sql-box { background: #1e2333; color: #c8d3f5; }
.r-box   { background: #0d3326; color: #9af0cc; }
.sql-box .label, .r-box .label {
  font-size: 10px; letter-spacing: .12em; text-transform: uppercase;
  margin-bottom: 6px; opacity: .6; font-family: sans-serif;
}
.sql-kw  { color: #89ddff; }
.sql-val { color: #f78c6c; }
.r-fn    { color: #2dd4bf; }
.r-arg   { color: #f5a623; }

/* ── Progress tracker ── */
.progress-section {
  background: #fff; border: 1px solid #e2dfd8; border-radius: 10px;
  padding: 16px 20px; margin: 1.5rem 0;
}
.progress-section h4 { margin: 0 0 10px; font-size: 13px; color: #6b6760; text-transform: uppercase; letter-spacing: .1em; }
.prog-items { display: flex; gap: 8px; flex-wrap: wrap; }
.prog-item {
  padding: 4px 12px; border-radius: 99px; font-size: 12px; font-weight: 500;
  border: 1px solid #e2dfd8; color: #9b9790; background: #f7f6f2;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = ""
)
```

```{r load-packages, echo=TRUE}
library(tidyverse)
```

---

## About this notebook

This notebook teaches **data manipulation with tidyverse** using agricultural data (2025). Each exercise mirrors a SQL command — so you can see both approaches side by side.

<div class="progress-section">
<h4>Exercises</h4>
<div class="prog-items">
  <span class="prog-item">01 · filter()</span>
  <span class="prog-item">02 · select()</span>
  <span class="prog-item">03 · arrange()</span>
  <span class="prog-item">04 · group_by() + summarise()</span>
  <span class="prog-item">05 · left_join()</span>
</div>
</div>

---

## The datasets

Two tables are available for all exercises. They are linked by **district_id**.

<div class="schema-wrap">
  <div class="schema-card">
    <div class="sc-title">districts</div>
    <table>
      <tr class="key-col"><td>district_id</td><td>text 🔑</td></tr>
      <tr><td>district</td><td>text</td></tr>
      <tr><td>population</td><td>integer</td></tr>
    </table>
  </div>
  <div class="schema-card">
    <div class="sc-title">agriculture</div>
    <table>
      <tr class="key-col"><td>district_id</td><td>text 🔑</td></tr>
      <tr><td>year</td><td>text</td></tr>
      <tr><td>crop</td><td>text</td></tr>
      <tr><td>area_ha</td><td>integer</td></tr>
      <tr><td>prod_kg</td><td>integer</td></tr>
      <tr><td>value_real</td><td>decimal</td></tr>
    </table>
  </div>
</div>

```{r load-data}
districts   <- read_csv("Data/districts.csv",   show_col_types = FALSE)
agriculture <- read_csv("Data/agriculture.csv", show_col_types = FALSE)
```

```{r preview-districts}
glimpse(districts)
```

```{r preview-agriculture}
glimpse(agriculture)
```

---

## tidyverse vs SQL — quick reference

Every SQL clause maps directly to a tidyverse verb:

| SQL | tidyverse | What it does |
|-----|-----------|--------------|
| `SELECT col1, col2` | `select(col1, col2)` | Keep only certain columns |
| `WHERE condition` | `filter(condition)` | Keep only rows matching a condition |
| `ORDER BY col DESC` | `arrange(desc(col))` | Sort rows |
| `GROUP BY col` | `group_by(col)` | Define groups for aggregation |
| `SUM(col) AS total` | `summarise(total = sum(col))` | Compute aggregates per group |
| `JOIN ... ON key` | `left_join(..., by = "key")` | Combine two tables |

The **pipe operator `%>%`** chains operations together — equivalent to nesting SQL clauses. Read it as *"then"*:

```r
agriculture %>%        # start with this table, THEN
  filter(crop == "Rice") %>%   # keep only Rice rows, THEN
  select(district_id, prod_kg) # keep only these columns
```

---

## Exercise 01 · `filter()` {#ex01}

*SQL equivalent: `WHERE`*

<div class="sql-compare">
<div class="sql-box">
<div class="label">SQL</div>
<span class="sql-kw">SELECT</span> * <span class="sql-kw">FROM</span> agriculture<br>
<span class="sql-kw">WHERE</span> crop = <span class="sql-val">'Maize'</span><br>
&nbsp;&nbsp;<span class="sql-kw">AND</span> prod_kg > <span class="sql-val">5000</span>;
</div>
<div class="r-box">
<div class="label">tidyverse</div>
agriculture <span class="r-fn">%>%</span><br>
&nbsp;&nbsp;<span class="r-fn">filter</span>(<span class="r-arg">crop == "Maize"</span>,<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<span class="r-arg">prod_kg > 5000</span>)
</div>
</div>

<div class="exercise">
<div class="ex-label">Exercise 01</div>
<p>From the <code>agriculture</code> table, keep only rows where the <strong>crop is "Maize"</strong> and <strong>prod_kg is greater than 5000</strong>. How many districts produced more than 5,000 kg of Maize?</p>
</div>

<div class="hint">
In <code>filter()</code>, use <code>==</code> (double equals) to test equality for text. Separate multiple conditions with a comma (which means AND).
</div>

```{r ex01, eval=FALSE}
# Write your answer here
agriculture %>%
  filter(___ == "___",
         ___ > ___)
```

```{r ex01-solution, echo=TRUE, code_folding='hide'}
# Solution — click to reveal
agriculture %>%
  filter(crop == "Maize",
         prod_kg > 5000)
```

<div class="solution">
The result should show districts where Maize production exceeded 5,000 kg.
</div>

---

## Exercise 02 · `select()` {#ex02}

*SQL equivalent: `SELECT col1, col2`*

<div class="sql-compare">
<div class="sql-box">
<div class="label">SQL</div>
<span class="sql-kw">SELECT</span> district_id, crop, value_real<br>
<span class="sql-kw">FROM</span> agriculture;
</div>
<div class="r-box">
<div class="label">tidyverse</div>
agriculture <span class="r-fn">%>%</span><br>
&nbsp;&nbsp;<span class="r-fn">select</span>(<span class="r-arg">district_id, crop, value_real</span>)
</div>
</div>

<div class="exercise">
<div class="ex-label">Exercise 02</div>
<p>From the <code>agriculture</code> table, keep only the columns <strong>district_id</strong>, <strong>crop</strong>, and <strong>value_real</strong>. The result should have 240 rows and exactly 3 columns.</p>
</div>

<div class="hint">
List column names separated by commas inside <code>select()</code>. No quotes needed for column names.
</div>

```{r ex02, eval=FALSE}
# Write your answer here
agriculture %>%
  select(___, ___, ___)
```

```{r ex02-solution, echo=TRUE, code_folding='hide'}
# Solution — click to reveal
agriculture %>%
  select(district_id, crop, value_real)
```

<div class="solution">
The result should have 240 rows and 3 columns: district_id, crop, value_real.
</div>

---

## Exercise 03 · `arrange()` {#ex03}

*SQL equivalent: `ORDER BY col DESC`*

<div class="sql-compare">
<div class="sql-box">
<div class="label">SQL</div>
<span class="sql-kw">SELECT</span> * <span class="sql-kw">FROM</span> agriculture<br>
<span class="sql-kw">ORDER BY</span> value_real <span class="sql-val">DESC</span>;
</div>
<div class="r-box">
<div class="label">tidyverse</div>
agriculture <span class="r-fn">%>%</span><br>
&nbsp;&nbsp;<span class="r-fn">arrange</span>(<span class="r-fn">desc</span>(<span class="r-arg">value_real</span>))
</div>
</div>

<div class="exercise">
<div class="ex-label">Exercise 03</div>
<p>Sort the <code>agriculture</code> table by <strong>value_real</strong> from highest to lowest. Which district and crop has the highest agricultural value?</p>
</div>

<div class="hint">
Use <code>arrange(desc(column_name))</code> for descending order. For ascending order you would just write <code>arrange(column_name)</code>.
</div>

```{r ex03, eval=FALSE}
# Write your answer here
agriculture %>%
  arrange(___(___))
```

```{r ex03-solution, echo=TRUE, code_folding='hide'}
# Solution — click to reveal
agriculture %>%
  arrange(desc(value_real))
```

<div class="solution">
The first row should be the district/crop combination with the highest value_real.
</div>

---

## Exercise 04 · `group_by()` + `summarise()` {#ex04}

*SQL equivalent: `GROUP BY` + `SUM()`*

<div class="sql-compare">
<div class="sql-box">
<div class="label">SQL</div>
<span class="sql-kw">SELECT</span> district_id,<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<span class="sql-kw">SUM</span>(prod_kg) <span class="sql-kw">AS</span> total_prod_kg<br>
<span class="sql-kw">FROM</span> agriculture<br>
<span class="sql-kw">GROUP BY</span> district_id;
</div>
<div class="r-box">
<div class="label">tidyverse</div>
agriculture <span class="r-fn">%>%</span><br>
&nbsp;&nbsp;<span class="r-fn">group_by</span>(<span class="r-arg">district_id</span>) <span class="r-fn">%>%</span><br>
&nbsp;&nbsp;<span class="r-fn">summarise</span>(<span class="r-arg">total_prod_kg = sum(prod_kg, na.rm = TRUE)</span>)
</div>
</div>

<div class="exercise">
<div class="ex-label">Exercise 04</div>
<p>For each <strong>district_id</strong>, calculate the <strong>total production</strong> in kg. Name the new column <strong>total_prod_kg</strong>. The result should have one row per district (80 rows).</p>
</div>

<div class="hint">
<code>group_by()</code> defines the groups, then <code>summarise()</code> computes one value per group. Use <code>na.rm = TRUE</code> inside <code>sum()</code> to ignore missing values.
</div>

```{r ex04, eval=FALSE}
# Write your answer here
agriculture %>%
  group_by(___) %>%
  summarise(total_prod_kg = ___(prod_kg, na.rm = TRUE))
```

```{r ex04-solution, echo=TRUE, code_folding='hide'}
# Solution — click to reveal
agriculture %>%
  group_by(district_id) %>%
  summarise(total_prod_kg = sum(prod_kg, na.rm = TRUE))
```

<div class="solution">
The result should have 80 rows — one per district — with columns district_id and total_prod_kg.
</div>

---

## Exercise 05 · `left_join()` {#ex05}

*SQL equivalent: `JOIN ... ON`*

<div class="sql-compare">
<div class="sql-box">
<div class="label">SQL</div>
<span class="sql-kw">SELECT</span> d.district, a.crop,<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;a.prod_kg, d.population<br>
<span class="sql-kw">FROM</span> agriculture a<br>
<span class="sql-kw">JOIN</span> districts d<br>
&nbsp;&nbsp;<span class="sql-kw">ON</span> a.district_id = d.district_id;
</div>
<div class="r-box">
<div class="label">tidyverse</div>
agriculture <span class="r-fn">%>%</span><br>
&nbsp;&nbsp;<span class="r-fn">left_join</span>(districts,<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<span class="r-arg">by = "district_id"</span>) <span class="r-fn">%>%</span><br>
&nbsp;&nbsp;<span class="r-fn">select</span>(<span class="r-arg">district, crop, prod_kg, population</span>)
</div>
</div>

<div class="exercise">
<div class="ex-label">Exercise 05</div>
<p>Join <code>agriculture</code> and <code>districts</code> using <strong>district_id</strong> as the key. Then select only the columns <strong>district</strong>, <strong>crop</strong>, <strong>prod_kg</strong>, and <strong>population</strong>. The result should have 240 rows.</p>
</div>

<div class="hint">
<code>left_join(table2, by = "key_column")</code> adds columns from <code>table2</code> to your table, matching on the key. Chain a <code>select()</code> afterwards to keep only the columns you need.
</div>

```{r ex05, eval=FALSE}
# Write your answer here
agriculture %>%
  left_join(___, by = "___") %>%
  select(___, ___, ___, ___)
```

```{r ex05-solution, echo=TRUE, code_folding='hide'}
# Solution — click to reveal
agriculture %>%
  left_join(districts, by = "district_id") %>%
  select(district, crop, prod_kg, population)
```

<div class="solution">
The result should have 240 rows and 4 columns: district, crop, prod_kg, population.
</div>

---

## Going further — combining verbs {#bonus}

Real analyses chain multiple verbs together. Try these on your own:

```{r bonus-1}
# Which districts have the highest total Bean production?
# (GROUP BY + filter + arrange + join to get district names)
agriculture %>%
  filter(crop == "Bean") %>%
  group_by(district_id) %>%
  summarise(total_prod_kg = sum(prod_kg, na.rm = TRUE)) %>%
  arrange(desc(total_prod_kg)) %>%
  left_join(districts, by = "district_id") %>%
  select(district, total_prod_kg) %>%
  head(10)
```

```{r bonus-2}
# Total value and production per crop across all districts
agriculture %>%
  group_by(crop) %>%
  summarise(
    total_prod_kg  = sum(prod_kg,    na.rm = TRUE),
    total_value    = sum(value_real, na.rm = TRUE),
    n_districts    = n_distinct(district_id)
  ) %>%
  arrange(desc(total_value))
```

```{r bonus-3}
# Simple bar chart — total production by crop
agriculture %>%
  group_by(crop) %>%
  summarise(total_prod_kg = sum(prod_kg, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(crop, total_prod_kg), y = total_prod_kg, fill = crop)) +
  geom_col(width = 0.6) +
  coord_flip() +
  scale_fill_manual(values = c(Bean = "#a78bfa", Maize = "#f5a623", Rice = "#3ecf8e")) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = "Total production by crop — 2025",
    x        = NULL,
    y        = "Production (kg)",
    fill     = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

## Summary — SQL → tidyverse cheatsheet

```{r cheatsheet, echo=FALSE}
tibble::tribble(
  ~SQL,                              ~tidyverse,                                         ~Notes,
  "SELECT col1, col2",               "select(col1, col2)",                               "Column names without quotes",
  "SELECT * (all columns)",          "no verb needed / select(everything())",            "Default is all columns",
  "WHERE col = 'x'",                 "filter(col == 'x')",                               "Use == not =",
  "WHERE col > 5 AND col2 = 'y'",    "filter(col > 5, col2 == 'y')",                    "Comma = AND",
  "WHERE col IN ('a','b')",          "filter(col %in% c('a','b'))",                      "Use %in% for lists",
  "ORDER BY col DESC",               "arrange(desc(col))",                               "ASC is default",
  "GROUP BY col",                    "group_by(col)",                                    "Always pair with summarise()",
  "SUM(col) AS total",               "summarise(total = sum(col, na.rm=TRUE))",          "na.rm avoids NA results",
  "COUNT(*) AS n",                   "summarise(n = n())",                               "n() counts rows",
  "JOIN t2 ON t1.key = t2.key",      "left_join(t2, by = 'key')",                       "Keeps all rows from left table",
  "SELECT DISTINCT col",             "distinct(col)",                                    "Remove duplicate values"
) %>%
  knitr::kable(col.names = c("SQL", "tidyverse", "Notes"))
```

---

*Dataset: Agricultural production,— 2025.  
Packages used: `tidyverse` (dplyr, ggplot2, readr).*
